---
author: "Austin Cherian"
date: "2024-11-24"
output: html_document
title: "Final Project pt. 2"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
#load necessary packages
library(enrichR)
library(openxlsx)
```

## R Markdown - Enrichment anlalysis using EnrichR R package

Summary
- Input gene lists created from T-Test / EdgeR group comparison analysis . Or an short list of genes of interest from your collaborator (Researcher/Clinician/)
- If needed, short list genes based on p-value or FDR value.
- Run EnrichR on short listed gene list

## Import T-test results so that we can short list further
```{r}
#folder that contains group comparison results
fileName <- "/Users/austincherian/HIDS_7003/finalProject/Part_1/code/output/20241123_TTest__Bladder mucosae surrounding cancer_(Comp).vs._Normal bladder mucosae_(Base).TTest.csv"

# read in group comparison results
result1 <- read.csv(file = fileName, 
                        header = T, stringsAsFactors = F, row.names = 1)
head(result1)
```

# Filtering/shortlisting example - 1
Short list results based in p-value cut off on FDR-adjusted p-values (FDR p-value <= 0.01)
```{r}
pValueCutOff <- 0.01

# Apply p.adjust to the Pvalue column of result1
adjustedPvalues <- p.adjust(result1$Pvalue, method = "BH")

# Find indices where adjusted p-values are <= 0.01
which1 <- adjustedPvalues <= pValueCutOff

# Summarize the number of genes passing the cutoff
table(which1) # 273 genes have FDR adjusted p-value less than the cutoff

# Subset the data to include only shortlisted genes
resultShort <- result1[which1, ]


```
## Export the short listed results for reference
```{r}
#length of short listed results
nrow(resultShort)

write.csv(x = resultShort, file = "/Users/austincherian/HIDS_7003/Class_07/Homework_07/output/TTest_results_shortlist2.csv")

head(resultShort)
```
## Clean gene names
Note - the gene names are in the form of "reporterid|genename". So need to split this.
```{r}
funcSplit <- function(rep_gene) {
    rep_gene_split <- unlist(strsplit(x = rep_gene, 
                                      split = "|", 
                                      fixed = TRUE))
    gene <- rep_gene_split[2]
    return(gene)
}
geneListSplit <- apply(X = as.matrix(row.names(resultShort)), 
                       MARGIN = 1, FUN = funcSplit )
head(geneListSplit) #cleaned gene names, 82 genes

#remove duplicates
geneListSplit1 <- unique(geneListSplit) #74 genes

# remove NA value
geneListFinal <- na.omit(geneListSplit1) #74 genes

head(geneListFinal)
```

```{r}
#print number of unique genes
length(geneListFinal)
```

```{r}
write.table(x = geneListFinal, 
          file = "/Users/austincherian/HIDS_7003/Class_07/Homework_07/output/shortListedUniqueGenes.tsv",
          quote = F, sep = "\t", row.names = FALSE, col.names = F)
```


## Load Databases for Enrichr R package , and check connection
```{r}
#checking if EnrichR website and packing are working
#dbs <- enrichR::listEnrichrDbs()  #total number of databases available = 200+ 

#testing if EnrichR package is working
testOutput <- enrichR::enrichr(genes = c("Runx1", "Gfi1", "Gfi1b", "Spi1", "Gata1", "Kdr"), databases = "KEGG_2021_Human")
head(testOutput[[1]])

#List of databases for which enrichment analysis will be run
dblist1 <- read.csv(file = "/Users/austincherian/HIDS_7003/finalProject/Part_2/code/input/2023-EnrichR-Databases.txt", 
                    header = F, stringsAsFactors = F)

head(dblist1)
```

## Call function to run Enrichment
Warning: genes must be a non-empty vector of gene names or a data.frame with genes and score.

```{r, message = FALSE, warning = FALSE}
#convert gene list to a dataframe
geneList_df <- as.data.frame(geneListFinal)

# set output file name
outputFileName1 <- paste("/Users/austincherian/HIDS_7003/finalProject/Part_2/output/BladderCancer", "_EnrichR.xlsx", sep="")

#Load R script into the environment
source(file = "/Users/austincherian/HIDS_7003/finalProject/Part_2/code/functionEnrichment.R") 

#call function to run Enrichment
functionEnrichment(dblist1, geneList_df, outputFileName1)

#NEED INTERNET CONNECTION
```