---
title: "Final Project"
author: "Austin Cherian"
output: html_document
date: "2024-11-23"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Analysis plan of action

Check uploaded docx file

## Step 0

### Clinical Data File Checks

Opened clinical data file in excel to get familiar with the data. Ran
the following checks:

-   Rows are patients
-   Columns are clinical attributes
-   There is one row per patient
-   Check/Notice the Number of rows and number of columns
-   All data must be de-identified

In Excel, all of the above checks were performed. The number of rows
(patients) was 233, and the number of columns (features) was 21.

### Gene Expression Data File Checks

-   Open gene expression data file in excel (if possible) to get
    familiar with the data. Run the following checks:
-   Columns are patients
-   Rows are features (genes in this case)
-   There is one column per patient
-   Check/Notice the Number of rows and number of columns
-   All data must be de-identified
-   It was given to us that gene expression data was normalized in log2
    scale and includes gene annotation.

In Excel, all of the above checks were performed. The number of columns
(patients) is 233, and the number of rows (genes) is 43148. The gene
expression data did include gene annotation.

## Step 1 - Read in data into R. Check if data was imported correctly

### Read clinical data into R

-   General Checks
    -   Rows are patients
    -   Columns are clinical attributes
    -   There is one row per patient
    -   How many patients are there (dimensions of the data frame)
-   Find the column which is your outcome of interest
-   What are the groups to compare?

### Read gene expression data into R

-   Read in cleaned processed molecular data into R
    -   Check dimensions of the data frame
    -   Rows are molecular features/variables (genes in our example)
    -   Columns are genes
    -   Rows are patients

#### 1a - Read in clinical data

```{r}

clinData <- read.csv(
  file = "/Users/austincherian/HIDS_7003/finalProject/code/input/BC_ClinData_233rows.csv", 
  sep = ",", 
  header = T, 
  stringsAsFactors = F,
  row.names = 1)

knitr::kable(head(clinData[,1:4],n = 5))
dim(clinData) #233 rows (patients), 21 columns (features)
```

#### 1b: Read in processed normalized gene expression data in log2 scale, includes gene annotation

```{r}
geneExp <- read.table(
  file = "/Users/austincherian/HIDS_7003/finalProject/code/input/BC_GeneExpData_withAnno_233.tsv",
  sep="\t",
  row.names = 1,
  header = T,
  stringsAsFactors = F,
  quote = "\"")
dim(geneExp) #Read in gene expression file - 43148 rows (genes), 233 columns (patients)
knitr::kable(geneExp[1:3, 1:4], caption =
"Processed gene expression data in log2 scale with gene annotation")
```

**The Goal is to compare which two groups ?**

-   Question: What is the outcome variable column from clinical data
    file ? It is the column in the clinical file that contains the
    baseline and comparison values
-   **Answer: The `PrimaryBladderCancerType` column has values “Normal
    bladder mucosae” (baseline group)** and **“Bladder mucosae
    surrounding cancer” (comparison group)**

```{r}
knitr::kable(table(clinData$PrimaryBladderCancerType)) 
```

## Step 2 - Clean/Filter data

**Filter clinical data**- Clinical data has one row per patient, which
is good. However, `PrimaryBladderCancerType` has values other than the
ones we wish to compare, so we need to filter those out.

```{r}
#logical vector that is True for baseline and comparison group values 
condition <- as.data.frame(clinData["PrimaryBladderCancerType"] == "Normal bladder mucosae" |clinData["PrimaryBladderCancerType"] == "Bladder mucosae surrounding cancer")
#filter clinData for values of interests
clinData_clean <- clinData[condition[,1],]
knitr::kable(table(clinData_clean$PrimaryBladderCancerType)) 
```

**Filter gene exp data**

```{r}
#How gene exp data looks before filter/cleaning
knitr::kable(geneExp[1:5, 1:4])
```

#### Subsetting Gene Expression Data

To analyze the data, we need to align the gene expression file with the
clinical data file by subsetting the gene expression data to include
only the patients present in the clinical data. This ensures that both
datasets correspond to the same set of patients.

How are the clinical and gene expression data connected? The connection
is established through patient identifiers (PIDs), which serve as the
unique keys for matching between the two files:

**Clinical data sample identifiers:** The GSMid column contains the
PIDs. **Gene expression data sample identifiers:** The column headers
represent the PIDs.

By matching these identifiers, we can subset the gene expression data
for the relevant patients in the clinical dataset.

```{r}
matchingSamples <- which(colnames(geneExp) %in% clinData_clean$GSMid) #68 IDs matching
subsetGeneExp <- geneExp[, matchingSamples] #43148 row features, 68 patient columns
```

## Step 3 - Identify the groups to be compared

### Identify the groups to be compared (Baseline and Comparison Grps)

```{r}
# Labels (row numbers) that can identify the baseline group patients
baselineGrpLabels <- which(clinData_clean$PrimaryBladderCancerType == "Normal bladder mucosae") 
head(baselineGrpLabels)

# Use the labels (row numbers) to subset baseline patients in clinical data file
clinBase <- clinData_clean[baselineGrpLabels, ] #10 samples

# Labels (row numbers) that can identify the comp group patients
compGrpLabels <- which(clinData_clean$PrimaryBladderCancerType == "Bladder mucosae surrounding cancer") 
head(compGrpLabels)

# Use the labels (row numbers) to subset comp patients in clinical data file
clinComp <- clinData[compGrpLabels, ] #58 samples
# Use the clinBase and clinComp objects to subset gene expression data
geneExpTumorBase <- subsetGeneExp[, clinBase$GSMid] #10 samples, 43148 features(rows)
geneExpTumorComp <- subsetGeneExp[, clinComp$GSMid] #58 samples, 43148 features(rows)
```

## Step 4: Sanity check

-   Saw if filtering of clinical data in R matches filtering of clinical
    data in excel
-   Saw if sample ids in clinical data match sample ids in gene exp data
    (if they don’t match it means your step 1 and/or 2 is wrong)
-   Verified seeing correct number of samples in baseline and comp
    groups
-   Exported the column names from gene expression data and checked to
    see if it contained only probe/gene names and no other garbage

```{r}
#See if sample ids in clinical data match sample ids in gene exp data
clinBase$GSMid == colnames(geneExpTumorBase)
clinComp$GSMid == colnames(geneExpTumorComp)

### Checking if the samples IDs baseline and comp groups are the same ---- you can put these in an excel
write.table(clinBase$GSMid , file = "~/HIDS_7003/finalProject/code/output/ClinBaseIDs.tsv", sep="\t", quote = F )
write.table(clinComp$GSMid , file = "~/HIDS_7003/finalProject/code/output/ClinCompIDs.tsv", sep="\t", quote = F )
write.table(colnames(geneExpTumorBase) ,
file = "~/HIDS_7003/finalProject/code/output/GeneExpBaseIDs.tsv",
sep="\t", quote = F )
write.table(colnames(geneExpTumorComp) ,
file = "~/HIDS_7003/finalProject/code/output/GeneExpCompIDs.tsv",
sep="\t", quote = F )
#Export the features from gene expression data
#Open this file and check that it contains only probe/gene names and no other garbage
write.table(rownames(geneExp),file = "~/HIDS_7003/finalProject/code/output/genExpFeatureIDsCheck.tsv", sep="\t", quote = F )
```

## Step 5: Prep data for T-test

-   Gene expression data must have features (genes in this case) as
    rows, and samples as columns.

-   Objects must be data frame

-   Numeric data only

### Checking to make sure data is a numeric data frame

```{r}
knitr::kable(head(geneExpTumorBase[1:5,1:4]))
knitr::kable(head(geneExpTumorComp[1:5,1:4]))
```

## Step 6- Call function for T-test

```{r}
source("~/HIDS_7003/finalProject/code/fnTTest.R")
#### Call T-test function
results1 = fnTTest(baseGroup = geneExpTumorBase,
compGroup = geneExpTumorComp,
testName = "20241123_TTest_",
baseGroupName = "Normal bladder mucosae",
compGroupName = "Bladder mucosae surrounding cancer",
folderName = "~/HIDS_7003/finalProject/code/output")
```
