---
title: "Average Stacked Bar using Cibersort Data"
output:
  html_document:
    df_print: paged
  html_notebook: default
---
```{r, include= FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))
library(knitr)
```


# Plot Average Stacked Bar using Cibersort Data - R Markdown

* Read in output from Cibersort online 
* Split into Normal and BC_Surr_Mucosa (pre-cancerous) groups. So you have one cibersort object for each group of interest. 
* For each group of interest
    + Plot average stacked for each group (this is one stacked bar with different colors , one color for each immune cell type)


## Load in data for Normal (Non-cancerous) and BC_Surr_Mucosa (Pre-cancerous) patient
```{r}
cibersortDataRelapse <- read.csv("/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/input/CIBERSORTx_BCSurrMucosa_Results.csv", header=TRUE, row.names=1)

cibersortDataRelapseFree <- read.csv("/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/input/CIBERSORTx_Normal_Results.csv", header=TRUE, row.names=1)
knitr::kable(cibersortDataRelapse[1:5, 1:4], caption = "cibersort data")
knitr::kable(cibersortDataRelapse[1:5, 1:4], caption = "cibersort data")
```


## Call function for each group to plot stacked bar graph
Output will be a set of graphs in the "outputFolderName" folder

```{r, echo=TRUE, results = 'show'}
source("functionStackedBarPlot.R")

#calling function for CRC Tumor Relapse
objectPlotRelapse <- functionStackedBarPlot(
    cibersortData = cibersortDataRelapse, #set Cibersort Data ojbect for plotting
    outputFileName = "Out_Ciber_CRCBCSurrMucosa_StackedBarPlot", #set output file name
    outputFolderName = "/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/output", #Set output folder name
    title1 = "CRC BC Surrounding Mucosa Stacked Bar Plot") #Change Title

### calling function for CRC Tumor No-relapse
objectPlotRelapseFree <- functionStackedBarPlot(
    cibersortData = cibersortDataRelapseFree, #set Cibersort Data ojbect for plotting
    outputFileName = "Out_Ciber_CRCNormal_StackedBarPlot", 
    #set output file name
    outputFolderName = "/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/output", #Set output folder name
    title1 = "CRC Normal Stacked Bar Plot") #Change Title

# Read and display saved plots from the output folder
BC_Surr_Mucosa_plot_path <- "/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/output/Out_Ciber_CRCBCSurrMucosa_StackedBarPlot.png"
Normal_plot_path <- "/Users/austincherian/HIDS_7003/finalProject/finalProject_part4_Cibersort/code/output/Out_Ciber_CRCNormal_StackedBarPlot.png"

# Insert images into the RMarkdown document
include_graphics(BC_Surr_Mucosa_plot_path)
include_graphics(Normal_plot_path)
```









